\begin{wbepi}{Mihaly Csikszentmihalyi (1990)}
Ideally, the end of extrinsically applied education should be the start of an education that is motivated intrinsically. At that point the goal of studying is no longer to make the grade, earn a diploma, and find a good job. Rather, it is to understand what is happening around one, to develop a personally meaningful sense of what one's experience is all about. From that will come the profound joy of the thinker.
\end{wbepi}

To those who continually encouraged me to love learning and thinking, without your support this work, and who I am today, would not exist.

I start with sincere thanks and appreciation for Joshua Coon, my research advisor. His guidance and mentorship these past years have truly made me a better scientist, writer, and communicator. With access to world-class science, the newest instruments, and a plethora of professional contacts, working in his research group has been extremely helpful in my development. The opportunities he provided me have truly propelled my career forward, and I am confident the training I received from him will continue to help me down the road.

In addition to Joshua's direct support, he has also assembled a great team of scientists and fellow graduate students that I have worked with. First and foremost would be our staff scientist Mike Westphall, who has provided incredible amounts of wisdom to me. If anyone deserved the title of advisor, it would be him. Alex Hebert and Christopher Rose receive special thanks for listening to all my crazy ideas and all their scientific advice they provided. To the other graduate students: Catie Mingoue, Anna Merrill, Greg Potts, Alicia Richards, Allison Balloon, Nick Riley, Nick Kwiecien, Arne Ulbrich, Tim Rhoads and Molly McDevitt, thank you for your support. Additionally, these former graduate students had a special part in my early years in the lab: Graeme McAlister, Doug Phanstiel, Violet Lee, Justin Brumbaugh, Aaron Ledvina, Amelia Peterson, Jason Russell, and Danielle Swaney. I especially want to thank Craig Wenger, who's expertise in computer related science is something I try to achieve.

Before my graduate work at Madison, there was my undergraduate education at Wheaton College, IL. And no two people, other than my parents, have had a bigger impact on my life than Daniel and Lisa Burden. It is under their mentorship and friendship that I first started a career in science and developed the skills to realize my dreams. Their willingness to individually train and guide students at Wheaton has surely touched more lives than just mine, but I am indebted to their commitment to me and my future. I strive to emulate their love of science, God, and for each other; they are truly people to model.

Without the strong and constant support of my family, all of this would be impossible. My parents have constantly provided for me, their support and love are without equal. Their encouragement to love learning, reading, the outdoors, and games have made me who I am today. 

Lastly, to my wife Rebecca, who has seen my graduate school journey first hand. I love you and the companionship we share, advice you provide, and the smile you bring me. Thank you for the life we had together and the future to come.